---
title: "delay"
author: "Rica Rebusit"
date: "2023-07-22"
output: html_document
---
I will be using this dataset from Kaggle -> [Click Me!](https://www.kaggle.com/datasets/ulrikthygepedersen/airlines-delay?resource=download&select=airlines_delay.csv) to predict if a flight will be delayed based on:

- **Flight**: Flight ID
- **Time**: Time of departure
- **Length**: Length of flight
- **Airline**: Airline ID
- **AirportFrom**: Which airport the flight flew from
- **AirportTo**: Which airport the flight flew to
- **DayOfWeek**: Day of the week of flight
- **Class**: Whether the flight was delayed depicted by (1) and (0) if not delayed

This dataset also has 539382 observations

# Loading Necessary Libraries and Importing dataset

```{r, warning=FALSE}
library(tidyverse)
library(randomForest)
```

```{r}
delay <- read.csv("~/Documents/delay-prediction/airlines_delay.csv")
```

# Checking the Data and EDA
Looking at the first 6 rows
```{r}
head(delay)
```

Checking the data types of each variable
```{r}
str(delay)
```

Looking at unique values of categorical variables
```{r}
unique(delay$Airline)
```
```{r}
unique(delay$AirportFrom)
```
```{r}
unique(delay$AirportTo)
```

Variable **Class** is a integer when we want it to be a factor in order to predict so convert **Class** to a factor
```{r}
delay$Class <- as.factor(delay$Class)
str(delay$Class)
```

Checking to see if there are NA values
```{r}
colSums(is.na(delay))
```

Dropping column **Flight** because there is no significance for the flight ID
```{r}
delay <- delay %>%
  select(-"Flight")
head(delay)
```

Now time to do some exploration by comparing **Airline** to other variables. First checking the count of delays in each airline
```{r}
ggplot(delay, aes(Airline, fill = Class)) + geom_bar(position="dodge") + theme_minimal() + scale_fill_brewer(palette = "Accent") + labs(title = "Airline Delay Count") + theme(plot.title = element_text(family = "Courier New", size = 20, face = "bold", hjust = 0.5))
```
It seems like WN has more delays
```{r}
count(filter(delay, Class == 1, Airline == "WN")) #WN has 65657 delays
```
Checking the count of delays
```{r}
ggplot(delay, aes(Class)) + geom_bar(fill = "Lightblue", color = "Blue") + theme_minimal() + labs(title = "Delay Count") + theme(plot.title = element_text(family = "Courier New", size = 20, face = "bold", hjust = 0.5))
```
More no delays than delays

Looking at time departure
```{r}
ggplot(delay, aes(Airline, Time)) + geom_boxplot(fill = "Lightblue", color = "Blue") + theme_minimal() + labs(title = "Airline Departure Time") + theme(plot.title = element_text(family = "Courier New", size = 20, face = "bold", hjust = 0.5))
```
All airlines seem to departure at the same time

```{r, warning=FALSE}
ggplot(delay, aes(Time, fill = Airline)) + geom_histogram(position="stack") + theme_minimal() + labs(title = "Departure Time Distribution") + theme(plot.title = element_text(family = "Courier New", size = 20, face = "bold", hjust = 0.5))
```

Looking at length of flight
```{r}
ggplot(delay, aes(Airline, Length)) + geom_boxplot(fill = "Lightblue", color = "Blue") + theme_minimal() + labs(title = "Airline Length Time") + theme(plot.title = element_text(family = "Courier New", size = 20, face = "bold", hjust = 0.5))
```
Seeing which the airline with max length and min length
```{r}
delay$Airline[which.max(delay$Length)] #CO
delay$Length[which.max(delay$Length)] #655
```
```{r}
head(filter(delay, Length == 655))
```
So airline CO departing from EWR flying to HNL has the greatest flight length

```{r}
delay$Airline[which.min(delay$Length)] #F9
delay$Length[which.min(delay$Length)] #0
```

```{r}
head(filter(delay, Length == 0))
```
Airline B6 also has flight length 0. Maybe the flights were cancelled and both looks like to be towards the weekend.

```{r, warning=FALSE}
ggplot(delay, aes(Length, fill = Airline)) + geom_histogram(position="stack") + theme_minimal() + labs(title = "Departure Length Distribution") + theme(plot.title = element_text(family = "Courier New", size = 20, face = "bold", hjust = 0.5))
```
Right skewed

Checking which day of the week is most popular for flights
```{r}
ggplot(delay, aes(DayOfWeek)) + geom_bar(fill = "Lightblue", color = "Blue") + theme_minimal() + labs(title = "Popular Days") + theme(plot.title = element_text(family = "Courier New", size = 20, face = "bold", hjust = 0.5))
```
During the middle of the weeks is most popular for flights

Now that we explored the data, dropped a column, and changed **Class** variable type, let's start to predict!

# Train and Test Data
Since we have a pretty large dataset, we can use a 80-20 split
```{r}
tra.tes <- sample(nrow(delay), 0.7*nrow(delay))
train <- delay[tra.tes, ]
test <- delay[-tra.tes, ]
```

# Logistic Regression
```{r}
#logistic <- glm(Class ~., data = train, family = binomial)
```

Dataset may be too large for my computer
